---
title: "Online sellers revisited worksheet"
author: "Joachim Vandekerckhove"
output:
  beamer_presentation:
    incremental: yes
header-includes:
  - \input{includes/beamerpheader.tex}
knit: (function(inputFile, encoding) {
    rmarkdown::render(inputFile, encoding = encoding, output_dir = "../pdf")
  })
---

## Set up environment

```{r setup, include=TRUE, echo=TRUE, message=FALSE}
library(rjags)
library(coda)
library(ggplot2)
library(dplyr)
library(readr)
```

## Data

Load the data and necessary colors.

```{r data}
k <- c(10, 48, 186,  75, 1)
n <- c(10, 50, 200, 100, 2)

nSellers <- length(k)

sellerColors <- c("#003366", "#B00000", "#FFD700",
                  "#8FBC8F", "#778899")
```

## JAGS Models

Create the JAGS model files for independent, aggregate, and hierarchical models and run MCMC sampling for each model.

```{r mcmc-sampling}
# Define parameters for MCMC
nChains  <-    8
nBurnin  <- 1000
nSamples <- 5000
nThin    <-    1

samplesList <- list()
```
##
```{r mcmc-sampling-independent, include=TRUE}
independent_model <- "
model{
  for (i in 1:nSellers){
    k[i] ~ dbin(theta[i], n[i])
    kPostpred[i] ~ dbin(theta[i], n[i])
    theta[i] ~ dunif(0, 1)
  }
}
"
modelFile <- "onlineSellersIndependent.jags"
writeLines(independent_model, con = modelFile)
```
##
```{r}
data_list <- list(k = k, n = n, nSellers = nSellers)

inits <- function() list(theta = runif(nSellers))

params <- c("theta", "kPostpred")
```
##
```{r}
jags_model <- jags.model(modelFile,
                         data     = data_list,
                         inits    = inits,
                         n.chains = nChains,
                         n.adapt  = nBurnin)
```
##
```{r}
update(jags_model, n.iter = nBurnin)
samples <- coda.samples(jags_model,
                        variable.names = params,
                        n.iter         = nSamples,
                        thin           = nThin)

samplesList[[modelFile]] <- samples
```

##
```{r mcmc-sampling-aggregate, include=TRUE}
aggregate_model <- "
model{
  for (i in 1:nSellers){
    k[i] ~ dbin(theta, n[i])
    kPostpred[i] ~ dbin(theta, n[i])
  }
  theta ~ dunif(0, 1)
}
"
modelFile <- "onlineSellersAggregate.jags"
writeLines(aggregate_model, con = modelFile)
```
##
```{r}
data_list <- list(k = k, n = n, nSellers = nSellers)

inits <- function() list(theta = runif(1))

params <- c("theta", "kPostpred")
```
##
```{r}
jags_model <- jags.model(modelFile,
                         data     = data_list,
                         inits    = inits,
                         n.chains = nChains,
                         n.adapt  = nBurnin)
```
##
```{r}
update(jags_model, n.iter = nBurnin)
samples <- coda.samples(jags_model,
                        variable.names = params,
                        n.iter         = nSamples,
                        thin           = nThin)

samplesList[[modelFile]] <- samples
```

##
```{r mcmc-sampling-hierarchical, include=TRUE}
hierarchical_model <- "
model{
  for (i in 1:nSellers){
    k[i] ~ dbin(theta[i], n[i])
    kPostpred[i] ~ dbin(theta[i], n[i])
    theta[i] ~ dnorm(mu, 1/sigma^2)T(0, 1)
  }
  mu ~ dunif(0, 1)
  sigma ~ dunif(0, 1)
}
"
modelFile <- "onlineSellersHierarchical.jags"
writeLines(hierarchical_model, con = modelFile)
```
##
```{r}
data_list <- list(k = k, n = n, nSellers = nSellers)

inits <- function() list(theta = runif(nSellers),
                         mu    = runif(1),
                         sigma = runif(1))

params <- c("theta", "kPostpred", "mu", "sigma")
```
##
```{r}
jags_model <- jags.model(modelFile,
                         data     = data_list,
                         inits    = inits,
                         n.chains = nChains,
                         n.adapt  = nBurnin)
```
##
```{r}
update(jags_model, n.iter = nBurnin)
samples <- coda.samples(jags_model,
                        variable.names = params,
                        n.iter         = nSamples,
                        thin           = nThin)

samplesList[[modelFile]] <- samples
```

## Inspect Results

Print the multivariate Gelman diagnostic for each model.

```{r inspect-results, results='asis'}
mpsrf <- list()
modelList <- c("onlineSellersIndependent.jags",
               "onlineSellersAggregate.jags",
               "onlineSellersHierarchical.jags")
for (modelName in modelList) {
  samples <- samplesList[[modelName]]
  gelman_diag <- gelman.diag(samples,
                             multivariate=TRUE)
  mpsrf[[modelName]] <- gelman_diag$mpsrf
}
gelman_df <- data.frame(Model = names(mpsrf),
                        MPSRF = unlist(mpsrf))
```

## Inspect Results

Print the multivariate Gelman diagnostic for each model.
```{r}
knitr::kable(gelman_df, row.names = FALSE,
             col.names = c("Model",
                           "Multivariate PSRF"))
```

## Analysis and Plots

Generate plots for posterior distributions and posterior predictive distributions.

```{r plot-posteriors, fig.width=10, fig.height=6, echo=FALSE}
p <- list()
for (modelName in modelList) {
  samples <- samplesList[[modelName]]
  chains <- as.matrix(samples)

  if (modelName == "onlineSellersIndependent.jags" ||
      modelName == "onlineSellersHierarchical.jags") {
    density_data <- data.frame()

    for (i in 1:nSellers) {
      density_vals <- density(
        chains[, paste0("theta[", i, "]")],
        from = 0, to = 1)
      density_data <- rbind(
        density_data,
        data.frame(x = density_vals$x,
                   y = density_vals$y,
                   Seller = paste("Seller", i)))
    }

    p[[modelName]] <- ggplot(density_data, aes(x = x, y = y, fill = Seller)) +
      geom_area(position = "identity", alpha = 0.6) +
      scale_fill_manual(values = sellerColors) +
      labs(title = paste(modelName),
           x = "Rate", y = "Density") +
      theme_minimal() +
      theme(legend.position = c(0.2, 0.9),
            legend.justification = "top",
            legend.key.size = unit(0.5, "lines"))
  } else if (modelName == "onlineSellersAggregate.jags") {
    density_vals <- density(chains[, "theta"], from = 0, to = 1)
    p[[modelName]] <- ggplot(data.frame(x = density_vals$x, y = density_vals$y), aes(x = x, y = y)) +
      geom_area(fill = sellerColors[length(sellerColors)], alpha = 0.6) +
      labs(title = "onlineSellersAggregate.jasp",
           x = "Rate", y = "Density") +
      theme_minimal() +
      theme(legend.position = c(0.2, 0.9),
            legend.justification = "top",
            legend.key.size = unit(0.5, "lines"))
  }
}
```
##
```{r, fig.width = 4, fig.height = 3, fig.align='center', echo=FALSE, message=FALSE}
print(p$onlineSellersIndependent.jags)
```
##
```{r, fig.width = 4, fig.height = 3, fig.align='center', echo=FALSE, message=FALSE}
print(p$onlineSellersAggregate.jags)
```
##
```{r, fig.width = 4, fig.height = 3, fig.align='center', echo=FALSE, message=FALSE}
print(p$onlineSellersHierarchical.jags)
```

##

```{r plot-joint-posterior, fig.width = 4, fig.height = 3, fig.align='center', echo=FALSE, message=FALSE}
samples <- samplesList[["onlineSellersHierarchical.jags"]]
chains <- as.matrix(samples)

mu <- chains[, "mu"]
sigma <- chains[, "sigma"]

ggplot(data.frame(mu = mu, sigma = sigma), aes(x = mu, y = sigma)) +
  geom_bin2d(bins = 50, aes(fill = after_stat(count))) +
  scale_fill_gradient(low = "white", high = sellerColors[1]) +
  labs(title = "Joint Posterior for Hierarchical Model", x = "Group Mean", y = "Group Standard Deviation") +
  theme_minimal()
```
