---
title: "If you're a Bayesian you can do anything that God forbids\n - Willem Heiser"
author: "Joachim Vandekerckhove"
output:
  beamer_presentation:
    incremental: yes
header-includes:
  - \input{includes/beamerpheader.tex}
knit: (function(inputFile, encoding) {
    rmarkdown::render(inputFile, encoding = encoding, output_dir = "../pdf")
  })
---

```{r setup, include=FALSE}
# Load necessary libraries
library(ggplot2)
library(rjags)
library(coda)
library(tibble)
library(dplyr)

# Set seed for reproducibility
set.seed(123)
```

## Things you can do if you're a Bayesian
  
 - Accumulate evidence for the absence of an effect
 - Draw valid conclusions with almost no data
 - Say things about means without knowing much about the basic units

## Order of operations

   In general, order of operations matters:
   $$f \circ g (x) \neq g \circ f (x)$$


## Estimating parameters is an operation

  In science, we often want to make statements about averages.\pause
  
  Estimating model parameters from data is an operation: $$\hat\theta = f(x)$$\pause
  
  Do we want the \emph{average model parameters of the data} or the \emph{model parameters of the average data}?
  
  $$\overline{f(x)} \neq f\left(\overline{x}\right)$$

## Insight learning

```{r fig.width = 4, fig.height = 3, fig.align='center', echo=FALSE, message=FALSE}
x <- 1:100
y <- ifelse(x < 50, 0, 1)
df1 <- data.frame(x = x, y = y)

p1 <- ggplot(df1, aes(x = x, y = y)) +
  geom_step() +
  labs(title = "One-shot learning, one participant", x = "Blocks", y = "Accuracy") +
  theme_minimal()

ggsave("includes/step1.eps", p1)
print(p1)
```


## Insight learning

```{r fig.width = 4, fig.height = 3, fig.align='center', echo=FALSE, message=FALSE}
df2 <- data.frame(x = rep(1:100, 10), 
                  y = unlist(lapply(1:10, function(i) ifelse(1:100 < sample(35:65, 1), 0, 1))), 
                  group = rep(1:10, each = 100))

p2 <- ggplot(df2, aes(x = x, y = y, group = group, color = as.factor(group))) +
  geom_step() +
  labs(title = "Ten participants", x = "Blocks", y = "Accuracy") +
  theme_minimal() +
  theme(legend.position = "none")

ggsave("includes/step2.eps", p2)
print(p2)
```


## Insight learning

```{r fig.width = 4, fig.height = 3, fig.align='center', echo=FALSE, message=FALSE}
curve <- function(x) {
  1 / (1 + exp(-0.1 * (x - 50)))
}
df3 <- data.frame(x = 1:100, y = curve(1:100))

p3 <- ggplot(df3, aes(x = x, y = y)) +
  geom_line() +
  labs(title = "Average of many participants", x = "Blocks", y = "Accuracy") +
  theme_minimal()
ggsave("includes/step3.eps", p3)
print(p3)
```

## Insight learning
  The ``average" learning curve looks nothing like the person-specific learning curve!
  
  \centering
  \includegraphics[scale=0.5]{includes/step1.eps}
  \includegraphics[scale=0.5]{includes/step3.eps}
  
  