---
title: "Online sellers worksheet"
author: "Joachim Vandekerckhove"
output:
  beamer_presentation:
    incremental: yes
header-includes:
  - \input{includes/beamerpheader.tex}
knit: (function(inputFile, encoding) {
    rmarkdown::render(inputFile, encoding = encoding, output_dir = "../pdf")
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rjags)
library(coda)
library(ggplot2)
library(tibble)
library(dplyr)
```

## Data

The data are online ratings data obtained from 3Blue1Brown (https://www.youtube.com/watch?v=8idr1WZ1A7Q).
They are very small, so we can enter them manually.


```{r data}
k <- c(10, 48, 186)
n <- c(10, 50, 200)
nSellers <- length(k)
```

## JAGS Model

The JAGS model is a direct translation of our model assumptions.

```{r model}
modelString <- "
model {
  for (i in 1:nSellers) {
    k[i] ~ dbin(theta[i], n[i])
    theta[i] ~ dunif(0, 1)
  }
}
"
jags_file <- "onlineSellers_solution.jags"
writeLines(modelString, con = jags_file)

dataList <- list(k = k, n = n, nSellers = nSellers)
```

## MCMC Sampling

To set up the MCMC sampler, begin by listing the parameter(s) of interest and writing an initial values generating function for them.

```{r mcmc-sampling-1}
# Parameters to monitor
params <- c("theta")

# Initial values
inits <- function() {
  list(theta = runif(nSellers))
}
```

## 

Compile and start adaptation for 1000 samples (per chain).

```{r mcmc-sampling-2}
# JAGS model
model <- jags.model(jags_file,
                    data     = dataList,
                    inits    =    inits,
                    n.chains =       8 ,
                    n.adapt  =    1000 )
```

## 

Do 1000 burn-in and then retain 5000 samples (per chain).
```{r mcmc-sampling-3}
# Burn-in
update(model, n.iter = 1000)

# Sampling
samples <- coda.samples(model,
                        variable.names = params,
                        n.iter         =  5000 )

# Convert to matrix for easier handling
samples_mx <- as.matrix(samples)
```

## Inspect the Results

The potential scale reduction factor is a standard diagnostic, it should be close to 1 and ideally less than 1.1.

```{r inspect-results-rhat}
print(gelman.diag(samples))
```

## 
Print some summary statistics.
```{r inspect-results-summary}
summary(samples)[1]$
  statistics[, c("Mean", "SD", "Time-series SE")] %>%
  print()
```

##
Visualize the posterior samples.
```{r analysis-0}
lineWidth <- 3
sellerLabel <- c("Seller 1: 10 / 10",
                 "Seller 2: 48 / 50",
                 "Seller 3: 186 / 200")

# Create data for plotting histograms
histogram_data <- lapply(1:nSellers, function(i) {
  data.frame(
    x = samples_mx[, paste0("theta[", i, "]")],
    Seller = sellerLabel[i]
  )
})

histogram_data <- bind_rows(histogram_data)
```

## 
Plot the histograms.
```{r analysis-1}
sellerColors <- c("#34568B", "#964F4C", "#F5E050")
d1 <- ggplot(histogram_data, aes(x = x, fill = Seller,
                                 color = Seller)) +
  geom_histogram(binwidth = 0.01, alpha = 0.6,
                 position = "identity") +
  scale_fill_manual(values = sellerColors) +
  scale_color_manual(values = sellerColors) +
  labs(x = "Rate", y = "Frequency",
       title = paste0("Posterior distributions of ",
                      "rate of positive ratings")) +
  theme_minimal() +
  theme(
    text = element_text(size = 20),
    legend.position = "top"
  )
```

## 
```{r analysis-f1}
print(d1)
```

```{r include=FALSE}
# Save the plot as PNG and PDF
ggsave("includes/onlineSellers_1.png", plot = d1, width = 11, height = 6.75, dpi = 300)
ggsave("includes/onlineSellers_1.pdf", plot = d1, width = 11, height = 6.75)
```
