---
title: "Signal detection theory"
author: "Joachim Vandekerckhove"
output:
  beamer_presentation:
    incremental: yes
header-includes:
  - \input{includes/beamerpheader.tex}
knit: (function(inputFile, encoding) {
    rmarkdown::render(inputFile, encoding = encoding, output_dir = "../pdf")
  })
---

## A psychological model: Signal detection theory
$${\mathcal{M}}_{sdt}: \begin{cases}
    \delta \sim N(1, 1)            & \beta \sim N(0, 1)\\
    \phi_h = \Phi( \delta/2-\beta) & \phi_f = \Phi(-\delta/2-\beta)\\
    h \sim B(\phi_h, n_s)          & f \sim B(\phi_f, n_n)\\
\end{cases}$$ \pause\small

    model {
        ## Complete this model
     }

## Signal detection theory ~ implementation
\small

     library(rjags)
     data <- list( h = 60 ,  sigtrials  = 100 ,
                   f = 11 ,  noistrials = 100 )
     modelString = "
         model {
             ## Complete this model
          }
     "

## Signal detection theory ~ implementation

     writeLines( modelString , con = "sdt.txt" )

     jagsModel = jags.model( file     = "sdt.txt" ,
                             data     =     data  ,
                             n.chains =        3  ,
                             n.adapt  =     1000  )
     set.seed(0)
     update( jagsModel , n.iter = 1000 )  # burn-in

     samples = coda.samples( jagsModel ,
                             variable.names = c("d", "b") ,
                             n.iter         =      10000 )


## Signal detection theory ~ results

```{r fig.width = 4.5, fig.height = 3.5, fig.align='center', message=FALSE, results = 'hide', echo=FALSE, jags}
library(rjags)

data <- list(h=60, sigtrials=100, f=11, noistrials=100)

modelString = "
    model {
        d ~ dnorm(1, 1)
        b ~ dnorm(0, 1)

        phih <- phi( d / 2 - b)
        phif <- phi(-d / 2 - b)

        h ~ dbin(phih, sigtrials)
        f ~ dbin(phif, noistrials)
     }
"
writeLines( modelString , con="sdt.txt")
parameters <- c("d",  "b")
jagsModel = jags.model( file="sdt.txt", data=data, n.chains=3, n.adapt=1000)
set.seed(0)
update(jagsModel, n.iter=1000)
samples = coda.samples(jagsModel, variable.names=parameters, n.iter=10000)

plot(samples)
```

## Signal detection theory ~ summary statistics
\small
```{r, echo=TRUE, summary}
summary(samples)$statistics

summary(samples)$quantiles
```

## Signal detection theory ~ convergence
\small
```{r, echo=TRUE, convergence}
effectiveSize(samples)

gelman.diag(samples)
```

